Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-15 2024pipit/section/30-Core.tex
bosch
eval: kind2

Streams $V$ are represented as a non-empty sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.

@@

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based, while the reactive semantics emphasises the finite-state streaming execution of the system.
We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.

>>>>

Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.

@@

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based, while the reactive semantics emphasises the finite-state streaming execution of the system.
We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital signal processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.

### 2024-01-15 2024pipit/section/50-Extraction.tex
eval: kind2

This translation to transition systems is verified to preserve the original semantics; however, it can duplicate work by requiring two calls to the step function.
To extract the above transition systems, we use \fstar{}'s normalisation-by-evaluation to generate step functions that update their mutable state in-place and which \lowstar{} can extract to C.
This normalisation strategy inlines the two step functions and is often able to remove the duplicate work.

Unfortunately, our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
We currently generate a monolithic step function for each Pipit program; an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

% \item staging: similarity to \cite{gallego2021w};

>>>>

This translation to transition systems is verified to preserve the original semantics.
To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
This specialisation results in a concrete transition system that fits in the \lowstar{}~\cite{protzenko2017verified} subset of \fstar{}, which can then be extracted to statically-allocated C code.

The translation for recursive streams described above calls the step function of the sub-stream twice, which can duplicate work.
The normalisation strategy used to partially-evaluate the translation inlines the two occurrences of the step function, and is often able to remove the duplicate work, but this removal is not guaranteed.
Our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
% We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

### 2024-01-15 2024pipit/section/60-Evaluation.tex
eval: kind2

Here, the local-time field denotes the time-since-boot in \emph{network time-units}, which matches the bitrate of the underlying network bus.
The mode-command is an optional field which indicates requests from the application to enter configuration or execution mode.
The transmission-status describes the status of the last transmission request and may be none, success, or various error conditions.
The bus-status describes whether the bus is currently idle, busy, or in an error state.
The two receive fields denote messages received from the bus; for application-specific messages the time-triggered logic only needs the message identifier.

@@

The enable-acks field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
The transmit fields denote whether to send a reference message or an application-specific message.
For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
To meet the schedule as closely as possible, the driver predicts the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.

The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
We generate real-time C code as described in \autoref{s:extraction}.
We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
The runtime of the driver logic is fairly stable: over 5,000 executions, the worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent, which our implementation satisfies.
For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$; however, as our runtime does not include the time required for the hardware-interop layer, we are not confident that our implementation will satisfy full-speed CAN bus.

Our code generation can be improved in a few ways.
A common optimisation in Lustre is to fuse consecutive if-statements with the same condition~\cite{bourke2017formally}; such an optimisation seems useful here, as our treatment of optional values introduces repeated unpacking and repacking.
Some form of array fusion~\cite{robinson2017machine} may also be useful for removing redundant array operations.
Our current extraction generates a transition-system with a step function which returns a tuple of the updated state and result.
Composing these step functions together requires repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
In the future, we plan to build on the current proofs to implement a more-sophisticated encoding that introduces less overhead.

% The Bosch M-TTCAN IP module limits to 64 trigger elements \cite{bosch2019mttcan}.

We have verified a simplification of the trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
We plan to verify the remainder of the TTCAN implementation and publish it separately.
Prior work formalising TTCAN has variously modeled the protocol itself~\cite{saha2007finite, pan2014modeling,li2018formal},
instances of the protocol~\cite{guo2020model},
and abstract models of TTCAN implementations~\cite{leen2006modeling}, but we are unaware of any prior work that has verified an \emph{executable} implementation of TTCAN.

Separately, Pipit has also been used to implement and verify a simple real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
The reservoir has a float switch to sense the water level and a solenoid to allow the intake of water.
The specification includes a simple model of the water reservoir and shows that the reservoir will not exceed the maximum level.

>>>>

Here, the local-time field denotes the time-since-boot in \emph{network time units}, which are based on the bitrate of the underlying network bus.
The mode-command is an optional field which indicates requests from the application to enter configuration or execution mode.
The transmission-status describes the status of the last transmission request and may be none, success, or various error conditions.
The bus-status describes whether the bus is currently idle, busy, or in an error state.
The two receive fields denote messages received from the bus; for application-specific messages the time-triggered logic only needs the message identifier.

@@

The enable-acks field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
The transmit fields denote whether to send a reference message or an application-specific message.
For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
To meet the schedule as closely as possible, the driver anticipates the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.

\subsection{Runtime}

The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
We generate real-time C code as described in \autoref{s:extraction}.
We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent.
For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$.
Further evaluation is required to ensure that the complete runtime including the hardware-interop layer is sufficient for full-speed CAN.

Our code generation can be improved in a few ways.
A common optimisation in Lustre is to fuse consecutive if-statements with the same condition~\cite{bourke2017formally}; such an optimisation seems useful here, as our treatment of optional values introduces repeated unpacking and repacking.
Some form of array fusion~\cite{robinson2017machine} may also be useful for removing redundant array operations.
Our current extraction generates a transition-system with a step function which returns a tuple of the updated state and result.
Composing these step functions together results in repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
In the future, we plan to build on the current proofs to implement a more-sophisticated encoding that introduces less overhead.

\subsection{Verification}

We have verified a simplified trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
For comparison, we implemented the same logic in the Kind2 model-checker~\cite{champion2016kind2}.
The restrictions placed on the triggers array --- that triggers are sorted by time-mark, that there must be an adequate time-gap between a trigger and its next-enabled, and that a trigger's time-mark must be greater-than-or-equal-to its index --- are naturally expressed with quantifiers.
The Kind2 model-checker includes experimental array and quantifier support~\cite{kind2userdoc}.
% but uses a custom syntax for arrays with no compiler support.
Due to the experimental nature of these features, we had to work around some limitations: for example, the use of arrays and quantifiers disables IC3-based invariant generation; quantified variables cannot be used in function calls; and the use of top-level constant arrays caused runtime errors that rendered most properties invalid~\cite{kind2024toparray}.

We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays up to size sixteen.
Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after \TODO{hours}.
In comparison, the Pipit version is verified \emph{for all sizes of arrays} and the checking took 5s of CPU time.
The M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
Both developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
We ran these experiments on a 2020 MacBook Air M1 with 16 gigabytes of RAM.

A critical simplification we made to the Kind2 implementation was to modify the trigger-enabled set to be a single cycle index.
In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
In our Pipit development, we can treat the definition of the cycle set abstractly.
However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.

% \begin{tabular}{rrr}
%   size & wall-clock & CPU time \\
%   4 & 8s & 11s \\
%   8 & 8s & 13s \\
%   16 & 125s & 276s \\
%   32 & - & - \\
% \end{tabular}

We plan to verify the remainder of the TTCAN implementation and publish it separately.
Prior work formalising TTCAN has variously modeled the protocol itself~\cite{saha2007finite, pan2014modeling,li2018formal},
instances of the protocol~\cite{guo2020model},
and abstract models of TTCAN implementations~\cite{leen2006modeling}, but we are unaware of any prior work that has verified an \emph{executable} implementation of TTCAN.

Separately, Pipit has also been used to implement and verify a real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
The reservoir has a float switch to sense the water level and a solenoid to allow the intake of water.
The specification includes a simple model of the water reservoir and shows that the reservoir does not exceed the maximum level under different failure-mode assumptions.

### 2024-01-15 2024pipit/section/70-Related.tex
causality etc
eval: kind2

% TODO arrays
\TODO{requires rewrite}

Although the FIR filter from XXX is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.

@@

The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.

@@

% cite also:
% hardware language in Agda \cite{harrison2021mechanized}
% Refinement types for Zélus \cite{chen2022synchronous}
% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}

@@

>>>>

Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
Compiling the triggers array with an unverified compiler such as Lustre V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
The Kind2 model-checker does include experimental array and quantifier support~\cite{kind2userdoc}, but uses a custom syntax for arrays with no compiler support.
This experimental support has some limitations: the use of arrays and quantifiers disables IC3-based invariant generation, quantified variables cannot be used in function calls, and top-level arrays are not supported~\cite{kind2024toparray}.

Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous} and JKind~\cite{gacek2018jkind} do not support quantifiers.
It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.

@@

The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}.
Recent work has used translation validation to show that the generated C code matches the high-level semantics~\cite{scott2023trustworthy}.
However, the model-checking suffers from the same semantic gap.

@@

% cite also:
% hardware language in Agda \cite{harrison2021mechanized}
% template-based invariant generation \cite{kahsai2011instantiation}

@@

% \subsection{Model-checking, counterexamples and invariant generation}
\subsection{Comparison to a combinator-based verification approach}
\TODO{compare to refinement types, verification conditions, etc here}

% Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.

% We could consider an alternative approach that was closer to an imperative program logic with preconditions and postconditions.
% For example, we could consider a richer type $\stream \tau~ \{ \rawbind{x}{\phi} \}$ which then the result, of type $\tau$, always satisfies stream postcondition $\phi$.
% We could imagine specifying count_when with such a type:

% \begin{tabbing}
%   @val@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\{ \top \} ~\NN~ \{ \rawbind{c}{0 \le c \le max} \}$\\
% \end{tabbing}

% An alternative would be to use refinement types \cite{chen2022synchronous} or some kind of program logic with pre- and post-conditions.

% We believe that such type-based approaches are useful for composing programs together, but 
%  the nested-recursive nature of Lustre programs makes such approaches cumbersome.
% In our system, the recursive combinator has type $@rec@: (\stream \alpha \to \stream \alpha) \to \stream \alpha$.
% Consider instead, if we had a refined stream $\stream \alpha \{ \psi \}$
